Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2014-01-11 doc/ttsp2013/ttsp_2013.tex
TTSP 2014 paper: *: Refactoring the introduction *: Cleaning out some of the preconditioning results *: Removing the solver verification results

\date{\today}
\title{Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation}
\begin{document}
\maketitle

@@

In this paper, we present Monte Carlo Synthetic Acceleration (MCSA) as
a linear solver technique for the simplified $P_N$ ($SP_N$)
discretization of the Boltzmann neutron transport equation. Using the
fully-formed linear operator for the transport problem, we explore
solutions to the $SP_N$ equations with MCSA using a challenging light
water reactor fuel assembly criticality calculation as the driving
problem. Several difficulties arise when applying MCSA to these
problems that were not observed in previous work where the method was
applied to more opaque thermal radiaton transport systems. In order to
effectively solve the $SP_N$ equations for the fuel assembly problem,
several algebraic preconditioners are studied within the context of
MCSA. For Jacobi-based preconditioners, convergence of the $SP_N$
system with MCSA is difficult if not impossible due to the lack of
diagonal dominance observed as a result the scattering generated by
light water reactor problems. To overcome these difficulties, we
utilize basic incomplete factorizations as preconditioners to achieve
convergence. Although the algebraic preconditioning strategy presented
here will not be practical from a performance implementation
standpoint, it does indicate the iterative performance of the method
when the operator spectrum is properly conditioned. Using the
preconditioners, MCSA performance is analyzed through comparison with
production Krylov solvers that would typically be used to solve
transport problems with an $SP_N$ discretization in terms of both
iterations to converge and CPU timing.

@@

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
reactor neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Brantley and Larsen \cite{brantley_simplified_2000}, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
or $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in neutron transport for fission reactor
analysis. The reduction in numerical complexity of current
deterministic solution methods using the $S_N$ approximation could
mean significant savings in both compute time and memory required. In
addition, the characteristics of the solution to the transport problem
for a steady state reactor core permit diffusion theory to be used; a
staple of the nuclear industry since its inception. Therefore, if
diffusion theory is applicable, then finer grained solutions that
capture more of the physics contained in the transport equation should
be possible with the $SP_N$ method. In doing so, we also expect from
the literature to obtain computed responses on the order of accuracy
we would expect from an appropriately discretized $S_N$ method at a
fraction of the cost.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simplified_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and algebraic
preconditioning methods.

This leads us to then research the applicability of Monte Carlo
Synthetic Acceleration methods as a possible solution method for the
$SP_N$ equations. In addition, solving the $SP_N$ equations in this
way also breaks away from the $S_N$ forms of parallelism where spatial
parallelism is achieved by an efficient parallel sweep, angular
efficiency achieved by pipe-lining, and energy parallelism achieved by
decoupling the groups. With the $SP_N$ equations as a full matrix
system, we now can parallelize the problem as prescribed by the Monte
Carlo algorithm, which may be significantly more scalable than current
$S_N$ transport practices. In this work we therefore apply Monte Carlo
Synthetic Acceleration to the $SP_N$ equations and research its
applicability, performance, and convergence properties in the context
of a light water reactor system.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}

Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \cite{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \cite{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.

In addition to the Monte Carlo solver parameters dictating the number
of histories and weight cutoff, the outer MCSA iterations also have
the following stopping criteria:
\begin{equation}
  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. As with any iterative
method, other stopping criteria using other vector norms could be
computed, however, for this work we will only use
Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
to tune in an MCSA implementation: the number of Monte Carlo histories
computed in the Neumann-Ulam solve during each MCSA iteration, the
weight cutoff for those histories, and the total MCSA convergence
tolerance as specified by $\epsilon$.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
I was thinking about dumping the SPN tech note derivation here. Its a
lot but it would be a good place to formally present the Exnihilo
discretization.

%%---------------------------------------------------------------------------%%
\section{Model Problem}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation\footnote{The operator will change if, for example, physics
  feedback through temperature or potentially burnup is
  considered. These additional physics will modify the cross sections
  used to assemble $\mathbf{M}$. The calculations presented here will
  consist of a single eigenvalue calculation with a static
  $\mathbf{M}$.} while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

%%---------------------------------------------------------------------------%%
\section{Preconditioned MCSA}

In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this by altering the eigenvalue spectrum of the
iteration matrix. It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underlying
Monte Carlo problem that will be solved with the Neumann-Ulam
method. We consider here the general left/right preconditioned method
as the left or right preconditioned methods can be inferred from its
formulation.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
part of a sequence for finding the substituted variable $\ve{u}$. We
do, however, gain the benefits of this general preconditioning by
building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator. In addition, for MCSA to apply to
increasingly difficult problems, more advanced preconditioning
techniques that require the generation of the composite operator may
be necessary for convergence.

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
A single energy group problem was first computed with
$SP_1$ discretization, effectively giving the one-speed neutron
diffusion system for the fuel assembly resulting in 20,088 degrees of
freedom in the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives
the residual infinity norm as a function of iteration for the MCSA
linear solve in the first eigenvalue iteration using 25,000 stochastic
histories at every iteration for the adjoint Neumann-Ulam solve.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.5in]{breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

\clearpage

>>>>

\date{\today} \title{A Preconditioned Monte Carlo
  Synthetic-Acceleration Method for the Simplified $P_N$
  Approximation}
\begin{document}
\maketitle

@@

In this paper, we present Monte Carlo Synthetic Acceleration (MCSA) as
a linear solver technique for the simplified $P_N$ ($SP_N$)
discretization of the Boltzmann neutron transport equation. Using the
fully-formed linear operator for the transport problem, we explore
solutions to the $SP_N$ equations with MCSA using a challenging light
water reactor fuel assembly criticality calculation as the driving
problem. Due to neutron scattering in the light water moderator,
several difficulties arise when applying MCSA to these problems that
were not observed in previous work where the method was applied to
more opaque thermal radiaton transport systems. In order to
effectively solve the $SP_N$ equations for the model problem, an
algebraic preconditioning strategy is developed for MCSA where we
utilize incomplete factorizations as preconditioners to achieve
convergence.  Using the preconditioners, MCSA performance is compared
with production impelementations of the preconditioned GMRES and
BiCGSTAB methods. Although the algebraic preconditioning strategy
presented in this work will not be practical from a high performance
implementation standpoint with results demonstrating poor time to
solution, it does indicate that the iterative performance of the
method is competitive with the preconditioned Krylov methods when
operator spectrum is properly conditioned.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel burn-up and therefore
reduction in generated waste, and high confidence in accident scenario
models. Modern high fidelity deterministic methods for large scale
problems are often variants on the discrete ordinates ($S_N$) method
\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
the $S_N$ method requires potentially trillions of unknown angular
flux moments to be computed to achieve good accuracy for the responses
of interest \cite{slaybaugh_acceleration_2011}.

As machines begin to operate at hundreds of petaflops peak performance
and beyond, trends toward reduced energy consumption will require
incredibly high levels of concurrency to achieve the desired
computation rates. The end result of these hardware changes is that
the larger number of low-powered processors will be prone to both soft
failures such as bit errors in floating point operations and hard
failures where the data owned by that processor cannot be
recovered. Because these failures are predicted to be common,
resilient solver technologies are required to overcome these events at
the application level. With solutions to deterministic transport
problems based on Monte Carlo techniques, such issues are potentially
alleviated by statistical arguments. In the case of soft failures,
isolated floating point errors in the Monte Carlo simulation are
absorbed within tally statistics while a hard failure is treated as a
high variance event where some portion of the Monte Carlo histories
lost.

Considering these physics and hardware issues, this goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. Using these numerical studies, we look
forward to parallelizing them with the goal that they may be leveraged
competitively on leadership-class computing platforms.

In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
synthetic acceleration (MCSA) method the thermal radiation diffusion
equation. For several transient problems we observed that MCSA
performed competetively with preconditioned Krylov methods. In a
continuation of these studies, we have applied the method light water
reactor problems discretized by the $SP_N$ equations. This system has
no time dependence, a large scattering component, and assymetry in the
linear operator, creating a set of problems complementary to the
transient, opaque and symmetric systems of our previous analyis.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the linear operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simplified_2013}. By fully forming
these equations and formulating them as a linear algebra problem
instead of using an explicit iterative method, we now have access to
modern advancements in computational linear algebra including Krylov
solvers for asymmetric systems and algebraic preconditioning methods.

The paper is organized as follows. First, we briefly introduce the
model $SP_N$ problem and describe the application of the MCSA
algorithm. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for neutron transport in light water reactors and subsquently develop
a new preconditioning scheme. We then apply this preconditioning
scheme to a light water reactor fuel assembly and compare the
performance of the preconditioned MCSA method to preconditioned Krylov
methods.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}):
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$. We apply the following change of
variables to yield the set of multigroup pseudo-moments
$\mathbb{U}_n$:
\begin{subequations}
  \begin{gather}
    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
where now the application of the fission matrix to the moment vectors
will be expanded into a set of block matrices in identical fashion to
the scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in
\cite{evans_simplified_2013}. It should be noted here that with
respect to a general MCSA scheme, the introduction of fission in the
system will only affect the source vector in the linear system for the
eigenvalue schemes used in this work. The linear operator, and
therefore overall MCSA performance will be dictated by streaming and
scattering as defined on the left-hand side.

\subsection{MCSA Solution of the $SP_N$ Equations}
\label{subsec:mcsa_solution}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \cite{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \cite{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. We refer the reader to
\cite{evans_monte_2012} for a more rigorous defintion of the method.

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

\subsection{Preconditioned MCSA}
\label{subsec:preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this by altering the eigenvalue spectrum of the
iteration matrix. It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underlying
Monte Carlo problem that will be solved with the Neumann-Ulam
method. We consider here the general left/right preconditioned method
as the left or right preconditioned methods can be inferred from its
formulation.

Several possible shortcomings of this preconditioning approach are
readily observed. First, the matrix-matrix multiplication operation
for sparse, parallel distributed matrices is significantly more
expensive than a matrix-vector multiplication operation. Second, each
preconditioner must be explicitly inverted, an operation in itself
that may be expensive and which prohibits the use of any
preconditioners which provide no mechanism to extract their
inverse. Third, for many modern preconditioning methods, this
inversion may yield dense matrices, destroying sparsity and further
impeding the performance of a matrix-matrix multiplication
operation. It is also interesting to note that the Monte Carlo problem
in the general left/right preconditioned scheme given by
Eq~(\ref{eq:left_right_correction}) is not fully left/right
preconditioned (meaning that we do not recover $\ve{x}$), but instead
part of a sequence for finding the substituted variable $\ve{u}$. We
do, however, gain the benefits of this general preconditioning by
building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator.

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Model Problem}
\label{sec:fuel_assembly}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Results}
In this section we present solutions to the fuel assembly problem
using MCSA preconditioning strategies.

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their similar diffusion-like form. A single energy
group problem was first computed with $SP_1$ discretization,
effectively giving the one-speed neutron diffusion system for the fuel
assembly resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.5in]{breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}
